Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-29 paper.tex
another pass through geometry section + rm some todos

In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
Specifically, in this paper we establish the exact radius and diameter of the space, [... -- Lena, please finish this including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.

@@

In \autocite{Gavryushkin2018-ol} an exact algorithm for computing the $\rnni$ graph has been introduced.
We implemented this algorithm and used it for computing this graph for trees on up to $7$ taxa.
\todo{LC: $7$ or $8$?}
\todo{Extend this section}

@@

\todo{LC: Work on the following proof! Is it worth introducing intervals for $\rnni$ moves for the proof of the following lemma?}

@@

    \todo{LC: revise this whole proof}

@@

    % Let $v:=\mrca_{\hat T}(C_i)$, let $C_{i+1}$ be the cluster induced by $v$, and let $w$ be the node inducing $C_i$.
    % We will distinguish the case that there is an edge $(v,w)$ in $T$ from the case that there is no such edge.
    %
    % At first we assume that there is no edge $(v,w)$, which means $C_i \not \subset C_{i+1}$.
    % In this case the two sets $C_i, C_{i+1}$ can exchange by a rank swap.
    % Since $C \subset C_{i+1}$, this rank swap decreases the rank of $\mrca(C)$ by one and there is no $\nni$ move that could move $\mrca(C)$ down.
    %
    % Let us now consider the case that $(v,w)$ is an edge in $T$.
    % The only move that could decrease the rank of $\mrca(C)$ is an $\nni$ move on the edge $(v,w)$.
    % Let $C_k$ and $C_j$ be the clusters induced by the children of $C_i$.
    % Then it is $C_k \cup C_j = C_i$.
    % There are two different $\nni$ moves possible on $(v,w)$.
    % $C_i$ can either be replaced by $C_j \cup (C_{i+1} \setminus C_i)$ or by $C_k \cup (C_{i+1} \setminus C_i)$.
    % To prove that there is always exactly one $\rnni$ moves that decreases the rank of $\mrca(C)$, we need to show that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
    %
    % Let $a \in C$.
    % Then it is $a \in C_{i+1} = mrca(C)$ and it follows $a \in C_{i+1} \setminus C_i$ or $a \in C_i$.
    % If $a \in C_{i+1} \setminus C_i$, $a$ is in the new set after the $\nni$ move.
    % So now we will assume that $a \in C_i$.
    % Then it is either $a \in C_j$ or $a \in C_k$.
    % Without loss of generality we assume that $a \in C_j$.
    %
    % We need to prove that if $a \in C_j$, there is no $b \in C$ with $b \in C_k$.
    % This makes sure that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
    % For contradiction we assume that $a \in C_j$ and $C \in S_k$.
    % We know that all sets that are subsets of $C$ already are at the same position in the current tree as in the destination tree $R$.
    % Therefore, we can conclude from $a \in C_j$ and $b \in C_k$ that $C_j \cup C_k = C$.
    % Hence, it holds $mrca(C) = C_i$.
    % This is a contradiction to our assumption that $C_{i+1} = mrca(C)$ in the current tree.
    % Therefore, it holds either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
    %
    % Concluding, there is always an $\rnni$ move on the edge $(v,w)$ for $mrca(C) = C_{i+1}, C \subsetneq C_{i+1}$ that decreases the rank of $mrca(C)$.

@@

With Lemma~\ref{lemma:mrca_move} we can conclude that $\findpath$ is a deterministic algorithm.
\todo{LC: move this up (before lemma1?)}
Since it computes a path between two arbitrary trees in $\rnni$, we can use the length of such a path as approximation for the distance between the trees.
A natural question is how well this estimates the actual distance in $\rnni$.
To answer this question for small trees, we implemented $\findpath$.
\todo{LC: github - link}
For computing the actual $\rnni$ distances we used the the algorithm for computing the $\rnni$ graph that was introduced in Section 3.3 of \autocite{Gavryushkin2018-ol}.
With our implementation of this algorithm
\todo{LC: gihub - link}
we can compute the graph for trees with up to $8$ taxa in reasonable time. \todo{LC: Shall we say more about the algorithm here? + running time (in hours/minutes) on which computer?}
We used
\todo{LC: How do we compare the distances?}

The running time of $\findpath$, implemented as described by the pseudocode in Algorithm~\ref{alg:find_path}, is quadratic in the number of taxa.
\todo{LC: Where do we mention the running time?}

@@

\todo{LC: move definitions of caterpillar stuff here}
\todo{LC: pseudo code}

@@

Let us now introduce a representation of caterpillar trees as lists of taxa.
This representation is basically a simplification of the cluster representation for caterpillar trees and works as follows.
For a caterpillar tree $T = [C_1, \ldots, C_{n-2}]$
\todo{LC: Do we need to mention that we refer to a tree by this notation, or do we need to say that this is the cluster representation of a tree?}
it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
Therefore, the list of taxa $L_T = [x_1, x_2, \ldots, x_n]$ with $x_1<x_2$ uniquely represents the caterpillar tree $T = [C_1, \ldots, C_{n-2}]$ with $C_i = \{x_1, \ldots, x_{i+1}\}$.
In this list representing a caterpillar tree the taxa are ordered non-decreasing, according to the ranks of their parents.
Note that the only two elements in such a list that have parents of equal rank are the first two elements $x_1$ and $x_2$.
That is why we order them in the list such that $x_1 < x_2$ to make the list representation of caterpillar trees unique.
We will call a pair of two taxa that share their parent \emph{cherry}.
We will use this list representation to show how one can adapt the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.

@@

Because all moves between caterpillar trees are $\nni$ moves, all pairs of taxa that are not in the same order in $L_T$ and $L_R$ need to swap at some point on any path between $L_T$ and $L_R$.
And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path.
\todo{LC: Do we need further explanation why $\csort$ gives a shortest caterpillar path?}

@@

The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
Within Section~\ref{section:caterpillar_convex} we will use $\csort$ for some proofs.
Therefore it is important to notice that at the end of each loop $i$ starting in Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort}, the last $i$ elements of $L_{T'}$ are the same as the ones in $L_R$.
This is a well known property of Bubble Sort that $\csort$ inherits.
\todo{LC: cite}

@@

The algorithm we are going to introduce here will be important for finding the radius of the $\rnni$ graph.
In Section~\ref{section:diameter} we will prove that it computes a tree with maximum distance from a given caterpillar tree, which implies that the radius of $\rnni$ equals its diameter.
Therefore, it is called $\mdtree$ (Maximum Distance Tree).
\todo{LC: References to the corresponding proofs?}

$\mdtree$ works as follows for an input tree $T$:
At first, all taxa of $T$ are sorted in a list $L$ such that the ranks of their parents are non-decreasing.
This representation is somehow similar to the list representation of caterpillar trees, but it is ambiguous since we do not specify in which order taxa appear in the list that share their parent.
If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$, in an arbitrary order, and $\parent(L[i])$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
$\mdtree$ builds the output tree $R$ as follows:
Initially, $R$ only consists of a cherry with taxa $L[0]$ and $L[1]$ such that the parent of these taxa has rank $n-1$.
All following taxa in $L$ are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.

@@

In this section we analyse shortest paths in $\rnni$.
Therefore, we will use the algorithms introduced in the previous section.
They enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph, before we discuss a conjecture, the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version.

Before we get into the details of these geometric properties of $\rnni$, we need to prove some statements that will be used a few times within this geometry section.
At first we in Lemma~\ref{lemma:diameter_bound} an upper bound of the \emph{diameter} $\Delta(\rnni)$ of the $\rnni$ graph, which is defined as $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$.

\begin{lemma}
$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
\label{lemma:diameter_bound}
\end{lemma}

\begin{proof}
Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C_i$, considered in Line~\ref{alg:find_path:line:cluster} of Algorithm~\ref{alg:find_path}, to be moved by $n-1-i$ $\rnni$ operations in $\hat{T}$.
Hence, the maximum length of such a path is bounded by $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

\begin{proof}
First observe that the rank of the internal node $\parent_T(n+1)$ can only be changed by performing a rank move that involves $\parent_T(n+1)$ or an $\nni$ move on an edge adjacent to $\parent_T(n+1)$.
Second observe that any $\rnni$ move can change the rank of $\parent_T(n+1)$ by at most one.
So we can follow that the rank of $\parent_T(n+1)$ can be decreased by at most one by an $\rnni$ move on $T$.

Let $p$ be a shortest path from tree $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
This new path $p{\big|}_n$ is then a path from $T{\big|}_n$ to $R{\big|}_n$.
The reason for this is that the only moves that are deleted from $p$ to receive $p{\big|}_n$ are those involving the parent of taxon $n+1$.
All other moves stay the same and can be performed equally on a path from $T{\big|}_n$ to $R{\big|}_n$ as on a path from $T$ to $R$, because the relations of all other internal nodes are the same between both pairs of trees.
Recall that $\delta$ is the difference in ranks between the parents of taxon $n+1$ in $T$ and $R$.
With the knowledge that an $\rnni$ move on $T$ can change the rank of $\parent_T(n+1)$ by at most one, we can conclude that $|p{\big|}_n| \leq |p| - \delta$.
Since $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n|$ and $|p| = d(T,R)$, the desired inequality follows.
\end{proof}

Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish the following Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.

@@

In this section we restrict our attention to the set of \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:algorithms}.
In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.

Therefore, we compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
In $\nni$, building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move resolves it.
This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
Hence a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
    \vspace{2pt}
    \todo{LC: Change caption}
	\caption{Paths between caterpillar trees $T$ and $R$: black -- shortest path in classical $\nni$ (ignoring ranks) of length $5$; red -- shortest path in $\rnni$ of length $6$; blue -- path in $\rnni$ which is the natural extension of the shortest path in $\nni$ but has length $8$.}
	\label{fig:NNI_vs_RNNI}
\end{figure}

@@

Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
Notice that by Lemma~\ref{lemma:diameter_bound} $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$.

@@

\begin{proof}
    Let $T$ and $R$ be caterpillar trees.
    Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
    This easily follows from algorithm $\csort$:
    In each loop $i=1, \ldots, n-2$ of the algorithm there can be at most $n-1-i$ exchanges of taxa.
    It follows that the maximum length of a path computed by this algorithm is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.

@@

    Induction step: $n \to n+1$\\
    Induction hypothesis: For caterpillar trees $T, R$ on $n$ taxa with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ it is $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
    Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
    Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
    Then each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
    In particular, taxon $n+1$ exchanges with a taxon in each of the $n+1$ loops in $\csort$.
    \todo{LC: Is this explanation sufficient?}
    We can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
    Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
    $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
    We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = n-1$.
    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
    By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
\end{proof}

@@

\begin{proof}
    We prove this Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
    Since we use backwards induction, the maximum distance between any two caterpillar trees, which is $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:diameter_bound}), is the base case.

@@

With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
Furthermore, we are able to show that the \emph{radius} of the $\rnni$ graph, which is defined as $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ for a graph $G$, equals its diameter.
At first we consider the diameter of the $\rnni$ graph.
By Lemma~\ref{lemma:diameter_bound} we already know an upper bound for the diameter.
We will now show that this is not only a bound, but the exact diameter of $\rnni$.

@@

%Do we need this?

% \begin{lemma}
% 	Let $T$ and $R$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
% 	There is a shortest paths from $T$ to $R$ that is a caterpillar path.
% 	\label{lemma:caterpillar_subset_convex}
% \end{lemma}
%
% \begin{figure}[h]
% 	\centering
% 	\includegraphics[width=0.3\textwidth]{label_internal_nodes}

>>>>

In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
Specifically, in this paper we establish the exact radius and diameter of the space, [... -- Lena, please finish this including the approximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.

@@

In Section 3.3 of \autocite{Gavryushkin2018-ol} an exact algorithm for computing the $\rnni$ graph has been introduced.
We implemented this algorithm and used it for computing this graph for trees on up to seven taxa.
\todo{LC: $7$ or $8$? - I'll need to check this}
Once the graph is computed, one can get distances by using algorithms such as Floyd-Warshall \autocite{Floyd1962-ew} or Dijkstra \autocite{Dijkstra1959-ph}.
However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we already run out of space when we want to save all pairwise distances for trees on seven taxa.
Therefore, we compute distances between certain pairs of trees and use the symmetry of $\rnni$ to draw conclusions about the remaining distances.

@@

@@

@@

@@

With Lemma~\ref{lemma:mrca_move} we can conclude that $\findpath$ is a deterministic algorithm.
Its running time is quadratic in the number of taxa.
Since it computes a path between two arbitrary trees in $\rnni$, we can use the length of such a path as approximation for the distance between the trees.
A natural question is how well this estimates the actual distance in $\rnni$.
To answer this question for small trees, we implemented $\findpath$.
\todo{LC: github - link}
For computing the actual $\rnni$ distances we used the the algorithm for computing the $\rnni$ graph as described in Section~\ref{section:alg_RNNI_graph}.
\todo{LC: How do we compare the distances?}

@@

@@

Let us now introduce a representation of caterpillar trees as lists of taxa.
This representation is basically a simplification of the cluster representation for caterpillar trees and works as follows.
For a caterpillar tree with cluster representation $T = [C_1, \ldots, C_{n-2}]$ it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
Therefore, the list of taxa $L_T = [x_1, x_2, \ldots, x_n]$ with $x_1<x_2$ uniquely represents the caterpillar tree $T = [C_1, \ldots, C_{n-2}]$ with $C_i = \{x_1, \ldots, x_{i+1}\}$.
In this list representing a caterpillar tree the taxa are ordered non-decreasing, according to the ranks of their parents.
Note that the only two elements in such a list that have parents of equal rank are the first two elements $x_1$ and $x_2$.
That is why we order them in the list such that $x_1 < x_2$ to make the list representation of caterpillar trees unique.
We will call a pair of two taxa that share their parent \emph{cherry}.
We will use this list representation to show how one can adapt the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.

@@

Because all moves between caterpillar trees are $\nni$ moves, all pairs of taxa that are not in the same order in $L_T$ and $L_R$ need to swap at some point on any path between $L_T$ and $L_R$.
And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path.

@@

The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
Within Section~\ref{section:caterpillar_convex} we will use $\csort$ for some proofs.
Therefore it is important to notice that at the end of each loop $i$ starting in Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort}, the last $i$ elements of $L_{T'}$ are the same as the ones in $L_R$.
This is a well known property of Bubble Sort that $\csort$ inherits from it.

@@

The algorithm we are going to introduce here will be important for finding the radius of the $\rnni$ graph.
In Section~\ref{section:diameter} we will prove that it computes a tree with maximum distance from a given caterpillar tree (Lemma~\ref{lemma:max_dist_caterpillar}), which implies that the radius of $\rnni$ equals its diameter (Corollary~\ref{corollary:radius}).
Therefore, it is called $\mdtree$ (Maximum Distance Tree).

$\mdtree$ works as follows for an input tree $T$:
At first, all taxa of $T$ are sorted in a list $L$ such that the ranks of their parents are non-decreasing.
This representation is somehow similar to the list representation of caterpillar trees, but it is ambiguous since we do not specify in which order taxa appear in the list that share their parent.
If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$, in an arbitrary order, and the parent of taxon $L[i]$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
$\mdtree$ builds the output tree $R$ as follows:
Initially, $R$ only consists of a cherry with taxa $L[0]$ and $L[1]$ such that the parent of these taxa has rank $n-1$.
All following taxa in $L$ are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.

@@

In this section we analyse shortest paths in $\rnni$.
Therefore, we will use the algorithms introduced in the previous section.
They enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph.
Interestingly, the \emph{diameter}, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, of $\rnni$ equals its \emph{radius} $rad(\rnni) =  \min\limits_T \max\limits_R d(T,R)$ in $\rnni$.
After proving this, we are going to consider the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version.

Before we get into the details of these geometric properties of $\rnni$, we need to prove some statements that will be used a few times within this geometry section.

\begin{lemma}
    It is $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
    \label{lemma:diameter_bound}
\end{lemma}

\begin{proof}
Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C_i$, considered in Line~\ref{alg:find_path:line:cluster} of Algorithm~\ref{alg:find_path}, to be moved by $n-1-i$ $\rnni$ operations in $T'$.
This can happen if each cluster $C_i$ has the root as $\mrca_{T'}(C_i)$.
Then the rank of this most recent common ancestor must be changed from $n-1$ to $i$, which requires $n-1-i$ operations, according to Lemma~\ref{lemma:mrca_move}.
Hence, the maximum length of such a path is bounded by $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

\begin{proof}
Let $T$ and $R$ be trees as described in the lemma.
First observe that the rank of the internal node $\parent_T(n+1)$ can only be changed by performing a rank move that involves $\parent_T(n+1)$ or an $\nni$ move on an edge adjacent to $\parent_T(n+1)$.
Second observe that any $\rnni$ move can change the rank of $\parent_T(n+1)$ by at most one.
So we can follow that an $\rnni$ on $T$ move can decrease the rank of $\parent_T(n+1)$ by at most one.

Let $p$ be a shortest path from $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
Then $p{\big|}_n$ is a path from $T{\big|}_n$ to $R{\big|}_n$.
The reason for this is that the only moves that are deleted from $p$ to receive $p{\big|}_n$ are those involving the parent of taxon $n+1$.
All other moves stay the same and can be performed equally on a path from $T{\big|}_n$ to $R{\big|}_n$ as on a path from $T$ to $R$, because the relations of all other internal nodes are the same between both pairs of trees.
Recall that $\delta$ is the difference in ranks between the parents of taxon $n+1$ in $T$ and $R$.
With the knowledge that an $\rnni$ move on $T$ can change the rank of $\parent_T(n+1)$ by at most one, we can conclude that $|p{\big|}_n| \leq |p| - \delta$.
Since $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n|$ and $|p| = d(T,R)$, the desired inequality follows.
\end{proof}

\todo{LC: We don't use the following Proposition anywhere in the paper. Do we want to keep it?}

Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.

@@

In this section we restrict our attention to the set of caterpillar trees, which we will represent throughout this section in the simplified list representation as introduced in Section~\ref{section:alg_csort}.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:algorithms}.
\todo{LC: Do we explain how we did this in the alg section? If not, we need to add an explanation here!}
In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.

Therefore, we compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
In $\nni$, building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move can resolve it.
This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
Hence a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
    \vspace{2pt}
	\caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
    The upper path, which is the $\rnni$ path that is the natural extension of the shortest $\nni$ path, is longer than a shortest $\rnni$ path as the one at the bottom.}
	\label{fig:NNI_vs_RNNI}
\end{figure}

@@

Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
Notice that $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.

@@

\begin{proof}
    Let $T$ and $R$ be caterpillar trees.
    Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
    This easily follows from algorithm $\csort$:
    In each loop $i=1, \ldots, n-2$ of Algorithm~\ref{alg:csort} there are at most $n-1-i$ exchanges of taxa.
    This is due to the fact that after loop number $i$ the last $i$ elements in the list representation of the current tree on $p$ are sorted correctly as in the destination tree.
    Hence there are no more exchanges of pairs of taxa needed within the last $i-1$ of the sequences in loop $i$, and so only $n-1-i$ swaps of elements might appear within the first $n-i$ elements of the list.
    It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.

@@

    Induction step: $n \to n+1$\\
    Induction hypothesis: For caterpillar trees $T, R$ on $n$ taxa with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ it is $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
    Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
    Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
    Then each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $T$ to $R$ exchanges.
    In particular, taxon $n+1$ exchanges with a taxon in each of the $n+1$ loops in $\csort$.
    \todo{LC: Is this explanation sufficient?}
    We can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
    Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
    $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

    We are now using this observation to prove that the distance between $T$ and $R$ equals $\frac{n(n-1)}{2}$.
    It is obvious that it gives an upper bound for the distance as it is $\Delta(\rnni_{n+1}) \leq \frac{n(n-1)}{2}$ for the $\rnni$ graph $\rnni_{n+1}$ on $n+1$ taxa by Lemma~\ref{lemma:diameter_bound}.
    So we now need to prove that $\frac{n(n-1)}{2}$ is a lower bound for the distance between $T$ and $R$ as well.
    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
    We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\pare nt_R(n+1))| = n-1$.
    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
    By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
\end{proof}

@@

\todo{LC: Work on the following proof, especially the figure and explanation of it}

\begin{proof}
    We prove this Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
    For this backwards induction the maximum caterpillar distance, which is $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:caterpillar_dist=diameter}), is the base case.

@@

With the results of Section~\ref{section:caterpillar_convex} we can easily get the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in Theorem 7 of \autocite{Gavryushkin2018-ol}.
At first we consider the diameter of the $\rnni$ graph.
Lemma~\ref{lemma:diameter_bound} already gives an upper bound for the diameter.
We will now show that this is not only a bound, but the exact diameter of $\rnni$.

@@

\subsection{Cluster Theorem}
\label{section:cluster_theorem}
